Cache Oblivious B-trees

### 2011-12-09 paper.tex
added my macros
Adding the first pass at the introduction

\newcommand{\Search}{\textsc{Search(k)}}
\newcommand{\Insert}{\textsc{Insert(k, v)}}
\newcommand{\Scan}{\textsc{Scan(a, b)}}
\newcommand{\Delete}{\textsc{Delete(k)}}

@@

Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

@@

@@

\subsection{CO vs CA}

@@

@@

>>>>

\newcommand{\Search}{\textsc{Search(k)}}
\newcommand{\Insert}{\textsc{Insert(k, v)}}
\newcommand{\Scan}{\textsc{Scan(a, b)}}
\newcommand{\Delete}{\textsc{Delete(k)}}
\newcommand{\Succ}{\textsc{Succ(k)}}
\newcommand{\Pred}{\textsc{Pred(k)}}
\newcommand{\Member}{\textsc{Member(k)}}
%\newcommand{\Range}{\textsc{Range-Query(k$_{i}$,k$_{j}$)}}

@@

@@

Most algorithms assume an idealized model of computation that consists of a
single CPU, which performs instructions, and memory, which is a flat array
that can be read and written. However, in reality memory is not a flat array
but rather a hierarchy from super fast on chip caches, to slower main memory,
and even all the way down to the disk. When a program requests data, the data
is found somewhere in the memory hierarchy and typically carried up the
hierarchy to the fastest memory possible layer when the CPU is able to access
it. Unfortunately, this fast memory often has small capacity and once it is
filled and new data is requests some old entries must be evicted. Bring data
from lower to higher levels in the memory hierarchy often takes a long time.
In fact, access times between the different layers of the hierarchy can differ
by many orders of magnitude and often dominate the overall runtime performance
of an algorithm. External memory algorithms are wary of this fact and aim to
solve problem by optimizing the number of memory transfers it performs rather
than the number of instructions it executes. They often try to exploit the
fact the memory hierarchies transfer data in blocks rather than individual
atomic units. Thus, external memory algorithms are designed to not only limit
the number of memory transfers but also preserve the locality of data so that
when a memory access is performed, its cost is amortized.

@@

Since External memory algorithms are primarily concerned with efficient access
to data, it is not surprise that most problems can be reduced to finding a
suitable data structure that efficiently implement certain operations. A well
known external memory data structure is the B-tree. At a high level, B-tree
work by making their nodes the same size as a memory block in the hierarchy.
Furthermore, the fanout in the tree (the number of children) is also
proportional to the block size as well. The B-tree incurs $O(log(N))$ memory
transfers for \Search{}, \Insert{}, and \Delete{} and is widely used in
databases and filesystems.

\subsection{DAM vs CO}

The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in
a two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}. However,
in reality computing architectures typically have memory hierarchies that are
much larger than two. The argument for the DAM model is that in practice only
one level in the memory hierarchy tends to dominate the performance and thus a
two-level memory model is sufficient for all intents and purposes. However,
this does mean that, theoretically, B-trees can begin to incur an increasing
number of memory transfers if the dominating level in the hierarchy changes.

Additionally, the B-tree expects to know the the block size in the memory
hierarchy in advanced. This means that the parameterized in code need to be
changed for every level in the memory hierarchy which can be very difficult
and error prone. Additionally, tuning this parameter may even be impossible in
a heterogeneous computing environment.

The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
of computation. In the CO model the algorithm / data structure has no
information about the memory hierarchy. It does not know anything about the
speed of any level or the block size. At a conceptual level, the CO model
allow algorithm designers to reason about the simple two-level model like DAM
but prove results for any arbitrary memory hierarchy. In short, algorithms
that perform well in a CO model should, in theory, perform well regardless of
the block size in a memory hierarchy and thus should gracefully scale to
multi-level hierarchies as well.

@@

In \cite{BenderDemainColton} Bender et al. introduce the cache-oblivious B-tree (CO B-tree). The
CO B-tree is similar to normal the B-tree except that it does not have the
block size of levels in the memory hierarchy as explicitly parameters.

The CO B-tree is able to achieve the optimal search bound of $\Theta(log(N))$
memory transfers, the optimal scan bound of $\Theta(N/B)$ memory transfers and
efficient insertion and deletion operations. The CO B-tree was a seminal data
structure that gave rise to many others in the future.

This paper provides a survey of these different data structures. We start by
analyzing the original CO B-tree and explain how it works. This includes a
detailed description of the packed memory array (PMA) and strongly
weight-balance binary search trees (SWSBT) that are integral to the CO
B-tree's performance. Following this discussion we provide an analysis of the
CO String B-tree,a CO B-tree with variable length keys, the CO Streaming
B-tree, a CO B-tree that improves the cost of insertion at the expense of
search, and the CO Concurrent B-tree (which allows multiple processes to
access the data at the same time).

@@

High level idea:

We want to have a B-tree that is cache oblivious. Okay, well, first we still
need a way to actually store our data. We should use a tree. But what type of
tree? Obviously it needs to be balanced but anything else? Yes, it needs to
maintain the property that the number of descendants are $O(d^h)$. Why? It has
to do with how we will ultimately layout the tree in memory, but we will get
to that later.

Okay, so, we find the tree that we want to use. Now what? Well, first of all,
we need to make sure that it supports all of the operations that we want while
maintain this key property. Lets start with searching. We will layout the tree
in a vEB layout. This is helpful for two reasons. First, it is a nice way to
store it in memory and second because it gives us a search bound for free. Now
we need to make sure that it supports all of the other operations while
maintain the key property. I can show you how it handles insertions and
delete. Bam!

Okay, so now, we have a data structure that holds the amount of memory that we
want and it has the property that we will use later. Great! Now what? Well, we
need to actually store this in memory some how. We said that we will use a vEB
layout but that is only half of the story. We want a data structure that
maintain the locality of as many elements as once while also keeping extra
space to allow for future inserts. This is seemingly contradictory but we can
do it!

How? we use a PMA. The main problem is what inserts and delete may cause
splits and merges which may cause huge parts and nodes of the tree to move
around. So we need to first show how it is able to handle splits and merges
efficiently in a cache oblivious manner. bam! done! Great!

Now what? We also have the problem of updating pointers. If the points are
close to one anther that is not a big deal. However, it is possible that the
nodes are really far apart. Close means they are in the same block and thus
only incur a single memory transfer. Far apart means they are more. To solve
this problem we use buffer nodes and that works in a weird way. However, this
only gives us a bad bound, can we do better? ****

Yes! We use indirection. Indirection allows us to use another array and this
makes out pointer updates much more efficient.

We can now perform inserts and deletions faster. Additionally, our scans are
super fast as well. Search have not changed at all because we still have our
vEB structure and thus we are okay. But, we can also do a little better. If we
are okay without having scans, we can improve inserts and deletion since we no
longer have to maintain our indirection array.

The important part is that our analysis above we never treat the block size as
a constant. It is always a variable in our analysis but we are able to achieve
the desired bounds. Thus, we have a cache-oblivious B-tree that does
everything that we want.
